A Triple Origin for Twin Blue Stragglers in Close Binaries

### 2019-04-08 BSS7782.tex
updated

We propose that twin blue stragglers (BSs) in compact binaries evolve
through a phase of mass transfer from a giant outer tertiary companion
on to the inner binary.  We apply this scenario to the twin-BS binary
WOCS~ID~7782 in the old open cluster NGC~188.  This binary has two
comparable-mass main-sequence stars in an almost circular ($e \aplt
0.1$) orbit of $\aplt 10$\,days.  Our theoretical arguments are
supported by simulations of an inner binary that accretes from a outer
Roche-lobe overfilling star using the Astrophysicsal Multipurpose
Software Environment. At least 80\,\% of the mass liberated by the
tertiary star goes through a circum-binary disk before it is accreted
by in the inner binary, causing the two stars to turn into blue
stragglers.  In order to acquire relatively stable phase of mass
transfer the donor should be about 1.4\,\MSun\, and over-fill it's
Roche lobe before ascending the asymptotic giant-branch.  The outer
star eventually turns into a 0.43 to 0.54\,\MSun\, white dwarf in a
relatively wide $\apgt 5.8\,yr$ orbit.  Although, the scenario is
generic, it requires some fine-tuning to achieve parameters comparable
to the observed twin WOCS~ID~7782.  This system is best reproduced
when starting with a $m_{\rm 3} = 1.4$ M$_{\odot}$ outer star in an
220 to 1100~day orbit around an inner binary composed of an
1.1\,\,MSun\, primary and a $m_{\rm 2}=0.7$ to $0.9$M$_{\odot}$
secondary star in an 8.6 to 24\,day orbit.  Based on our simulations
and theoretical arguments we predict that twin blue-stragglers that
formed through mass transfer from a Roche-lobe over-filling outer star
are generally comparable in mass, have aligned spins which again are
algined with the orbit of the tertiary white-dwarf star.  If the two
inner stars ware initially unequal in mass the less massive star will
have accreted more material from the tertiary star, which therefore is
more enhanced in CNO-processed material.

@@

Adopting a mass for the tertiary star of $m_3 = 1.4$\,\MSun, we can
constrain the initial parameters for the inner binary as well as the
orbit of the outer star after mass transfer. We first calculate the
stellar radius as a function of core mass. In
figure\,\ref{fig:tertiarymass_vs_size} we present this relation
calculated using the {\tt SeBa} stellar evolution code
\citep{1996A&A...309..179P} as the dark blue curve.  The interruption
in this curve, around a core mass of $m_{\rm core} \sim 0.5$\,\MSun\,
is a result of the evolution along the horizontal branch, where the
core of the star continues to grow but the radius actually shrinks.
Roche-lobe overflow in this phase is not expected to happen, because
it would already have happened in an earlier evolutionary state of the
donor star, when it was bigger.

@@

The orbit of the inner binary expands in these cases as a result of
accretion onto the inner two stars. In all three cases for the
1.4\,\MSun\, donor presented in
figure~\ref{fig:mass_vs_semimajor_axis} the inner orbit expands at
about the same rate. Consequently, the inner binaries that start with
$a = 0.15$\,au and $a=0.20$\,au eventually become dynamically
unstable.  The binary with an initial separation of $0.10$\,au expands
to reach a separation of about 0.126--0.145\,au for final masses for
the inner two stars of 1.4\,\MSun, which is consistent with the
observed twin BS 7782. In our simulations the eccentricity of the
inner binary grows to about $e \simeq 0.0028$.

With the accretion of mass, both stars in the inner binary also
accrete angular momentum.  By the end of the simulation the spins of
the two BSs are aligned along the orbital angular momentum axis with
an angle of $90.0^\circ$ for the primary star and $93.4^\circ$ for the
secondary star with respect to the argument of pericenter of the inner
orbit.  This supports our naive prediction that the spin angular
momenta of the two stars in the inner binary should be more or less
aligned post-mass transfer, due to the non-negligible amount of mass
accreted. By the end of the simulations the spin of the primary is
about 50.5 rotations per day, and 41.5 rotations per day for the
secondary star.  Such high spin rates immediately post-mass transfer
are supported by other work (see, for example,
\citet{2014MNRAS.438.1909D}), but the twin blue stragglers in
WOCS~ID~7782 are not observed to be spinning that fast
\cite{2018ApJ...869L..29L}.  Rapdily spinning stars may slow due to,
for example, magnetic braking, bringing them closer to the actually
observed spin rates \cite{2018arXiv181202181L}.

@@

\item For the predicted outer tertiary orbit, the initial orbital
  period should lie between 220 days $\lesssim$ P$_{\rm out}$
  $\lesssim$ 1100 days, assuming initial masses for the inner binary
  components of $m_{\rm 1} = 1.1$ M$_{\odot}$ and $m_{\rm 2} = 0.9$
  M$_{\odot}$ and an initial outer tertiary mass of $m_{\rm 3} = 1.4
  $M$_{\odot}$. The final orbital period of the white dwarf
    around the binary blue straggler should exceeds the initial
    orbit, but be smaller than $\sim 4100$\,days.

@@

  We further validated this statement by performing an
    additional series of calculations in which we vary the mass of the
    tertiary star in the initial triple from 0.5\,\MSun\, to
    0.7\,\MSun\, and 0.9\,\MSun. In
    fig.\,\ref{fig:mass_ratio_evolution_for_14MSun} we present the
    evolution of the normalized mass ratio in these binaries. with
    these calculations we demonstrate that a low-mass ratio initially
    tends to evolve towards an equal mass ratio.

  \begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_mass_ratio_evolution_for_14MSun.pdf}
  \caption{Evolution of the mass ratio for initial triples with an
    inner orbital separation of 0.1\,au (orange dotted curve) and
    0.2\,au (all other curves). The initial primary mass was
    1.4\,\MSun\, overfilling it's Roche lobe at a radius of
    100\,\RSun. the companion masses are $m_2= 0.9$\,\MSun\, and the
    mass of the tertiary is indicated in the legend. The dotted black
    line indicates the required mass-ratio evolution in order to
    eventually reach an equal mass-ratio blue-straggler binary.
\label{fig:mass_ratio_evolution_for_14MSun}}
\end{figure}

@@

Finally, we emphasize that the choice for the initial mass of the
outer tertiary may be rather critical.  Mass transfer in our proposed
scenario proceeds from the most massive tertiary to a binary of lower
total mass. This may result in an unstable phase of mass transfer, in
particular if the donor has a convective envelope
\citep[e.g.][]{2009pfer.book.....M}. A radiative envelope of the donor
ensures that the mass transfer will be maximally conservative, such
that the accretion stream will be maximally stable, accreting at a
stable and roughly constant rate \citep[e.g.][]{1991ApJS...76...55I}.
This stability regime may also be of interest for explaining very
massive twins, of $\apgt 20$\,\MSun\, which could be promising sources
for gravitational wave detectors once both twins evolve to a binary
black hole \citep{2016MNRAS.460.3545D}.

@@

With an original outer star of mass $\sim 1.4$\,\MSun, the inner orbit
tends to expand. This eventually leads to a dynamically unstable
system resulting either in a collision or in the ejection of
(probably) the lowest mass star. This evolution could result in a
single ejected BS, with the other BS left in a relatively close and
eccentric orbit with a WD (the left-over core of the tertiary star).
Such a close BS-WD binary would be hard to explain in another way.
%% These ``imposter'' BS-WD binaries would in principle mimic what is
%% expected theoretically for BSs formed from mass transfer in binary
%% stars.
If such a dynamical instability engages relatively late in the
mass-tranfer phase, the white dwarf (maybe with a little left-over
envelope) is expected to be ejected. This would lead to a relatively
wide twin blue-straggler binary and a single low-mass white dwarf.

When we adopt an inner orbit of $0.10$\,au\, the expansion eventually
matches the observed orbital separation (i.e., $0.13$\,au) of the
observed twin BS 7782 and the observed masses of the two stars of
about 1.4\,\MSun.

>>>>

We propose that twin blue stragglers (BSs) in compact binaries evolve
through a phase of mass transfer from a giant outer tertiary companion
on to the inner binary.  We apply this scenario to the twin-BS binary
WOCS~ID~7782 in the old open cluster NGC~188.  This binary has two
comparable-mass main-sequence stars in an almost circular ($e \aplt
0.1$) orbit of $\aplt 10$\,days.  Our theoretical arguments are
supported by simulations of an inner binary that accretes from an outer
Roche-lobe overfilling star using the Astrophysical Multipurpose
Software Environment. At least 80\,\% of the mass liberated by the
tertiary star goes through a circum-binary disk before it is accreted
by in the inner binary, causing the two stars to turn into blue
stragglers.  In order to acquire a relatively stable phase of mass
transfer the donor should be about 1.4\,\MSun\, and over-fill it's
Roche lobe before ascending the asymptotic giant-branch.  The outer
star eventually turns into a 0.43 to 0.54\,\MSun\, white dwarf in a
relatively wide $\apgt 5.8\,yr$ orbit.  Although, the scenario is
generic, it requires some fine-tuning to achieve parameters comparable
to the observed twin WOCS~ID~7782.  This system is best reproduced
when starting with a $m_{\rm 3} = 1.4$ M$_{\odot}$ outer star in an
220 to 1100~day orbit around an inner binary composed of an
1.1\,\MSun\, primary and a $m_{\rm 2}=0.7$ to $0.9$M$_{\odot}$
secondary star in an 8.6 to 24\,day orbit.  Based on our simulations
and theoretical arguments we predict that twin BSs that
formed through mass transfer from a Roche-lobe over-filling outer tertiary
are generally comparable in mass and have aligned spins, which are in turn 
aligned with the orbit of the tertiary white-dwarf star.  If the two
inner stars were initially unequal in mass the less massive star will
have accreted more material from the tertiary star, and should therefore be 
more enhanced in CNO-processed material.

@@

Adopting a mass for the tertiary star of $m_3 = 1.4$\,\MSun, we can
constrain the initial parameters for the inner binary as well as the
orbit of the outer star after mass transfer. We first calculate the
stellar radius as a function of core mass. In
figure\,\ref{fig:tertiarymass_vs_size} we present this relation
calculated using the {\tt SeBa} stellar evolution code
\citep{1996A&A...309..179P} as the dark blue curve.  The interruption
in this curve, around a core mass of $m_{\rm core} \sim 0.5$\,\MSun\,
is a result of the evolution along the horizontal branch, where the
core of the star continues to grow but the stellar radius actually shrinks.
Roche-lobe overflow in this phase is not expected to happen, because
it would already have happened in an earlier evolutionary state of the
donor star, when it was bigger.

@@

The orbit of the inner binary expands in these cases as a result of
accretion onto the inner two stars. In all three cases for the
1.4\,\MSun\, donor presented in
figure~\ref{fig:mass_vs_semimajor_axis} the inner orbit expands at
about the same rate. Consequently, the inner binaries that start with
$a = 0.15$\,au and $a=0.20$\,au eventually become dynamically
unstable.  The binary with an initial separation of $0.10$\,au expands
to reach a separation of about 0.126--0.145\,au for final masses for
the inner two stars of 1.4\,\MSun, which is consistent with the
observed twin BS WOCS 7782. In our simulations the eccentricity of the
inner binary grows to about $e \simeq 0.0028$.

With the accretion of mass, both stars in the inner binary also
accrete angular momentum.  By the end of the simulation the spins of
the two BSs are aligned along the orbital angular momentum axis with
an angle of $90.0^\circ$ for the primary star and $93.4^\circ$ for the
secondary star with respect to the argument of pericenter of the inner
orbit.  This supports our naive prediction that the spin angular
momenta of the two stars in the inner binary should be more or less
aligned post-mass transfer, due to the non-negligible amount of mass
accreted. By the end of the simulations the spin of the primary is
about 50.5 rotations per day, and 41.5 rotations per day for the
secondary star.  Such high spin rates immediately post-mass transfer
are supported by other work (see, for example,
\citet{2014MNRAS.438.1909D}), but the twin blue stragglers in
WOCS~ID~7782 are not observed to be spinning that fast
\cite{2018ApJ...869L..29L}.  Rapidly spinning stars may slow due to,
for example, magnetic braking, bringing them closer to the actually
observed spin rates \cite{2018arXiv181202181L}.

@@

\item For the predicted outer tertiary orbit, the initial orbital
  period should lie between 220 days $\lesssim$ P$_{\rm out}$
  $\lesssim$ 1100 days, assuming initial masses for the inner binary
  components of $m_{\rm 1} = 1.1$ M$_{\odot}$ and $m_{\rm 2} = 0.9$
  M$_{\odot}$ and an initial outer tertiary mass of $m_{\rm 3} = 1.4
  $M$_{\odot}$. The final orbital period of the white dwarf
    around the binary blue straggler should exceed the initial
    orbit, but be smaller than $\sim 4100$\,days.

@@

    We further validated this statement by performing an additional
    series of calculations in which we vary the mass of the tertiary
    star in the initial triple from 0.5\,\MSun\, to 0.7\,\MSun\, and
    0.9\,\MSun. In fig.\,\ref{fig:mass_ratio_evolution_for_14MSun} we
    present the evolution of the normalized mass ratio in these
    binaries. With these calculations we demonstrate that a low-mass
    ratio initially tends to evolve towards an equal mass ratio.

  \begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_mass_ratio_evolution_for_14MSun.pdf}
  \caption{Evolution of the mass ratio for initial triples with an
    inner orbital separation of 0.1\,au (orange dotted curve) and
    0.2\,au (all other curves). The initial primary mass was
    1.4\,\MSun\, overfilling it's Roche lobe at a radius of
    100\,\RSun. The companion masses are $m_2= 0.9$\,\MSun\, and the
    mass of the tertiary are indicated in the legend. The dotted black
    line indicates the required mass-ratio evolution in order to
    eventually reach an equal mass-ratio blue straggler binary.
\label{fig:mass_ratio_evolution_for_14MSun}}
\end{figure}

@@

Finally, we emphasize that the choice for the initial mass of the
outer tertiary may be rather critical.  Mass transfer in our proposed
scenario proceeds from the most massive tertiary to a binary of lower
mass. This may result in an unstable phase of mass transfer, in
particular if the donor has a convective envelope
\citep[e.g.][]{2009pfer.book.....M}. A radiative envelope of the donor
ensures that the mass transfer will be maximally conservative, such
that the accretion stream will be maximally stable, accreting at a
stable and roughly constant rate \citep[e.g.][]{1991ApJS...76...55I}.
This stability regime may also be of interest for explaining very
massive twins, of $\apgt 20$\,\MSun\, which could be promising sources
for gravitational wave detectors once both twins evolve to a binary
black hole \citep{2016MNRAS.460.3545D}.

@@

With an original outer star of mass $\sim 1.4$\,\MSun, the inner orbit
tends to expand. This eventually leads to a dynamically unstable
system resulting either in a collision or in the ejection of
(probably) the lowest mass star. This evolution could result in a
single ejected BS, with the other BS left in a relatively close and
eccentric orbit with a WD (the left-over core of the tertiary star).
As discussed in Section~\ref{intro}, BS-WD binaries are predicted to be the products of the binary mass transfer hypothesis (ignoring a common envelope phase) for BS formation. However, this mechanism tends to predict wide orbits, which is consistent with the observed BS-WD systems in NGC 188.  
%% These ``imposter'' BS-WD binaries would in principle mimic what is
%% expected theoretically for BSs formed from mass transfer in binary
%% stars.
If, on the other hand, such a dynamical instability engages relatively late in the
mass-tranfer phase, the white dwarf (maybe with a little left-over
envelope) is expected to be ejected. This would lead to a relatively
wide twin blue-straggler binary and a single low-mass white dwarf.

When we adopt an inner orbit of $0.10$\,au\, the expansion eventually
matches the observed orbital separation (i.e., $0.13$\,au) of the
observed twin BS WOCS 7782 and the observed masses of the two stars of
about 1.4\,\MSun.
